Cellular Automata and Computational Universality

### 2018-09-18 paper/paper.tex
Oops last was old version
Paper updoots

\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{amsmath}
\usepackage{url}
\usepackage{diagbox}
\usepackage{enumitem}
\usepackage{braket}

@@

    \subsection{Universality of Game of Life}
    A Turing Machine may compute a function by starting with the input to the
    function on the tape and halting with the output of the function on the
    tape. A function $f : \Sigma \ast \rightarrow \Sigma \ast$ is computable if
    some Turing Machine $M$ halts, for every input $w$, with just the result
    $f(w)$ on the tape \cite{Computable}. Thus, any model we can devise which
    could simulate the run of a Turing Machine on some input would be capable of
    computing all the functions that the Turing Machine could, and hence such a
    model would be Turing complete.

@@

    \begin{figure}[h]
        \includegraphics{}
        \caption{}
        \label{}
    \end{figure}

\section{More cellular automata}
    \subsection{WireWorld}
    \subsubsection{Computational Universality}
    \subsection{Rule 110}
    \subsubsection{Computational Universality}
    \subsection{Von Neumann's automaton}
    \subsection{Fractran}
    \subsection{Langton's Ant}
\section{Applications}
    \subsection{Cryptography}
\section{Firing Squad Synchronisation Problem}
\section{Majority Problem}

>>>>

\usepackage[margin=1in]{geometry}

\usepackage{graphicx}
\graphicspath{{figs/}}

\usepackage{subfig}
\usepackage{amsmath}
\usepackage{url}
\usepackage{diagbox}
\usepackage{enumitem}
\usepackage{braket}

@@

    \subsection{Universality in the Conway's Game of Life}
    A Turing Machine may compute a function by starting with the input to the
    function on the tape and halting with the output of the function on the
    tape. A function $f : \Sigma \ast \rightarrow \Sigma \ast$ is computable if
    some Turing Machine $M$ halts, for every input $w$, with just the result
    $f(w)$ on the tape \cite{Computable}. So for each computable function, there
    is a Turing Machine $T$ such that running $T$ on input $x$ always halts and
    outputs the correct answer (i.e., ends with the correct answer on the tape).
    Thus, any model we can devise which could simulate the run of a Turing
    Machine on some input would be capable of computing all the functions that
    the Turing Machine could, that is, all computable functions, and hence such
    a model would be Turing complete.

    Processor instruction sets may be fully implemented using just a NOT gate
    and one of either AND or OR in order to compute all binary logic. All that
    remains to obtain Turing Completeness is access to infinite, random access
    memory. Of course, given the finite nature of computers, this is impossible,
    but as long as the memory is as large as is required by the Turing Machine,
    this will suffice. Memory can be implemented in logic circuits using D-type
    flip flops, which can be done with four NAND gates. Thus, if this can be
    implemented using a cellular automaton, that automaton will be a universal
    computer.

@@

    \begin{figure}[h]
        \centering
        \includegraphics[width=5in]{golTM.png}
        \caption{Section of a Turing Machine implemented in Conway's Game of
        Life}
        \label{fig:golTM}
    \end{figure}

\section{More cellular automata}
    \subsection{WireWorld}
    WireWorld is a two-dimensional cellular automaton first proposed by Brian
    Silverman in 1987. The ruleset is relatively simpleset, with four distinct
    states:
    \begin{enumerate}
        \item empty
        \item conductor
        \item electron head
        \item electron tail
    \end{enumerate}

    As the state names may betray, WireWorld is particularly good at simulating
    electronics. Like Game of Life, WireWorld uses a Moore neighbourhoodin its
    calculations, as opposed to the more restricted Von Neumann
    neighbourhood.The rules are as follows:
    \begin{itemize}
        \item $empty \rightarrow empty$
        \item $head \rightarrow tail$
        \item $tail \rightarrow conductor$
        \item $conductor \rightarrow head \text{ if exactly 2 neighbours are
            $head$, else } conductor$
    \end{itemize}

    \begin{figure}[h]%
        \centering \subfloat[OR and XOR gates]{{\includegraphics[width=2in]{wireworld_gates.png}}}%
        \qquad
        \subfloat[D-type flip flop]{{\includegraphics[width=2in]{wireworld_flipflop.png}}}%
        \caption{All the structures required for universal computation implemented in WireWorld}%
        \label{fig:nature_fractals}%
    \end{figure}

    \subsubsection{Computational Universality}
    As with Game of Life, WireWorld can be used to implement the logic gates
    required for Turing Completeness, and is inherently more suited to doing so.
    A full computer has been constructed in WireWorld, which calculates prime
    numbers \cite{WW_TM}. Langton's Ant, a cellular automaton that will be
    discussed shortly, has also been implemented in WireWorld \cite{WW_LA}.

    \subsection{Rule 110}
    \subsubsection{Computational Universality}
    The key to Rule 110's computational universality is its ability to simulate
    a ``tag system'', which is universal, using gliders inherent to the
    automaton \cite{Rule110_universal}. The details are somewhat beyond the
    scope and level of this report, but Cook first proves that a tag system that
    removes two symbols at each stage is universal by compiling a two-state
    Turing Machine program. After this, he shows that such a tag system may be
    implemented using these gliders in Rule 110, and consequently shows that
    this cellular automaton is capable of universal computation. Some such
    gliders used by Cook in his proof of Rule 110's universality are shown in
    Fig. \ref{fig:110_gliders}.

    \begin{figure}[h]
        \centering
        \includegraphics[width=4in]{rule110_gliders.png}
        \caption{Gliders that interact by a) translation and b) forming a new
        structure}
        \label{fig:110_gliders}
    \end{figure}

    Due to Rule 110's simplicity, it suggests naturally occurring physical
    systems may also be capable of universality (systems similar to the way the
    shapes formed on the cone snail's shell, for example). Therefore, many of
    the properties of such systems would be undecidable, for which it is
    impossible to find closed-form mathematical solutions.

    \subsection{Von Neumann's automaton}
    Cellular automata can trace their roots back to the work done in the late
    1940s by Stanislaw Ulam and John von Neumann. Von Neumann wanted to
    ultimately answer the question of whether a machine was capable of
    replicating itself. We are used to automata creating structures much simpler
    than themselves, for example a factory that produces bolts: presumably, one
    section of the machine would cast the metal, while another would make the
    thread. He was concerned whether there was some form of ``extra-mechanical
    magic'' to self-reproduction \cite{CA_origins}.

    By 1952, von Neumann had completed his goal of engineering an automaton
    capable of recreating itself. It used a total of 29 states, and each cell's
    state in the next generation is calculated by considering its von Neumann
    neighbourhood in the current one. A summary of the states is as follows:

    \begin{enumerate}
        \item one ground state, $U$
        \item eight transition states,
            $S,S_0,S_{00},S_{000},S_{01},S_1,S_{10},S_{11}$
        \item four confluent states, $C_{00}, C_{01}, C_{10}, C{11}$
        \item eight ordinary transmission states in directions North, East,
            South, and West
            \begin{itemize}
                \item North-quiescent, North-excited
                \item East-quiescent, East-excited
                \item South-quiescent, South-excited
                \item West-quiescent, West-excited
            \end{itemize}
        \item eight special transmission states in the same directions
            \begin{itemize}
                \item North-quiescent, North-excited
                \item East-quiescent, East-excited
                \item South-quiescent, South-excited
                \item West-quiescent, West-excited
            \end{itemize}
    \end{enumerate}

    The ruleset can be broken down into four sections, concerning those for
    construction, destruction, transmission states, and confluent states. They
    are extensive and a full summary can be found at \cite{VN_rules}.

    \subsection{Fractran}
    \subsection{Langton's Ant}
\section{Applications}
    \subsection{Cryptography}
\section{Firing Squad Synchronisation Problem}
    \subsection{The Problem}
    Imagine a group of soldiers all part of a firing squad, with a general at
    one end. Firing squads would aim to have all of its soldiers fire at the
    same time so that no one soldier could be sure he had delivered the fatal
    wound to the condemned. The soldiers of the firing squad can be modelled as
    cells in a cellular automaton, and so the Firing Squad Synchronisation
    Problem, or FSSP, is concerned with ensuring that all states end up in a
    ``firing'' state at some point, at the same point in time. At each time
    step, a soldier may change what he is doing based on what him and his two
    immediate neighbours are doing (so the two soldiers on the ends only take
    into account their own current action and that of his single neighbour).
    Assume all soldiers start off in the same inactive state, also called the
    quiescent state. The question begins when the general initially gives the
    command to fire: assuming there is a fixed number of states, regardless of
    how many soldiers are in the line, how can the states be defined such that
    all soldiers are in the firing state at the same time?

\section{Majority Problem}
